---
title: "Exercise - Specification"
subtitle: "Why use particular functional transformations for the variables?"
published-title: "Date"
date: "2024-10-02"
self-contained: true
engine: knitr
---

```{r packages_hidden}
#| echo: false
#| message: false

library(webexercises)
```

::: {.callout-important title="Instructions"}
This page describe the instructions and some interactive snippets. You will have to run most of your code on `RStudio`, on your own computer.
:::

First, let's load useful packages. We need the `tidyverse` package but also to install the `gapminder` package (it contains the dataset we are interested in).

```{r packages}
#| message: false
#| code-fold: true

# install.packages("gapminder")

library(tidyverse)
library(gapminder)
```

## Exploring the data set

Load the package by calling `gapminder::gapminder`. What does each variable represent?

## Ploting the relationship

Plot a scatter plot of `lifeExp` as a function of `gdpPercap`.

```{r}
#plot
```

Does the relationship look linear? `r mcq(c("Yes", answer = "No"))`

What if you plot the log of `gdpPercap`? `r mcq(c(answer = "Yes", "No"))`. Of `lifeExp` `r mcq(c("Yes", answer = "No"))`
